---
title: "Circular–circular regression on the torus"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5.4, dpi = 130
)
```

A circular–circular (C–C) regression has an angle on both sides: the
response $\theta$ and the predictor $\varphi$ each live on the circle. In
`circlss` this is just a circular-response GAM whose covariate enters
through one of mgcv's *cyclic* spline bases, so the fitted curves close
seamlessly at $\pm\pi$:

$$\theta_i \sim \mathrm{vM}\big(\mu(\varphi_i),\ \kappa(\varphi_i)\big),
\qquad
\mu = 2\arctan(\eta_1),\quad \kappa = e^{\eta_2},$$

with $\eta_1 = f_1(\varphi)$ and $\eta_2 = f_2(\varphi)$ two cyclic
smooths. The natural picture of C–C data is a **torus**: the predictor
angle runs around the ring, the response angle around the tube, and the
fitted mean direction $\hat\mu(\varphi)$ is a closed curve winding over
the surface.

## Simulate data

```{r}
library(mgcv)
library(circlss)

# von Mises deviates, Best & Fisher (1979) rejection sampling
rvm <- function(n, mu, kappa) {
  mu <- rep_len(mu, n); kappa <- rep_len(kappa, n); out <- numeric(n)
  for (i in seq_len(n)) {
    k <- kappa[i]
    a <- 1 + sqrt(1 + 4 * k * k); b <- (a - sqrt(2 * a)) / (2 * k)
    r <- (1 + b * b) / (2 * b)
    repeat {
      z <- cos(pi * runif(1)); f <- (1 + r * z) / (r + z)
      cc <- k * (r - f); u2 <- runif(1)
      if (cc * (2 - cc) - u2 > 0 || log(cc / u2) + 1 - cc >= 0) {
        out[i] <- sign(runif(1) - 0.5) * acos(max(min(f, 1), -1)) + mu[i]
        break
      }
    }
  }
  atan2(sin(out), cos(out))
}

set.seed(20260612)
n <- 200
phi <- runif(n, -pi, pi)
mu_true <- 2 * atan(1.6 * sin(phi))
kappa_true <- exp(1.4 + 0.6 * cos(phi))
dat <- data.frame(theta = rvm(n, mu_true, kappa_true), phi = phi)
```

## Fit

Both distribution parameters get their own cyclic smooth; the knots pin
the period to $(-\pi, \pi]$. Smoothing parameters are selected by full
Newton REML (the family implements log-likelihood derivatives to fourth
order).

```{r}
b <- gam(list(theta ~ s(phi, bs = "cc", k = 10),
                    ~ s(phi, bs = "cc", k = 10)),
         family = vmlss(), data = dat, method = "REML",
         knots = list(phi = c(-pi, pi)))
summary(b)
```

Everything mgcv provides for general families works as usual — here are
the two estimated smooths, on the link scale with credible bands
($\eta_1$: tan-half location; $\eta_2$: log concentration):

```{r smooths, fig.height = 4}
plot(b, pages = 1, scheme = 1)
```

## Flat view

The fitted mean direction $\hat\mu(\varphi)$ on the response scale, with
the data. With this link the mean stays inside $(-\pi, \pi)$, the
Fisher–Lee branch:

```{r flat}
phig <- seq(-pi, pi, length.out = 400)
pred <- predict(b, newdata = data.frame(phi = phig), type = "response")
mu_hat <- pred[, 1]

op <- par(mar = c(4, 4, 1, 1))
plot(dat$phi, dat$theta, pch = 19, col = adjustcolor("#2c5aa0", 0.5),
     cex = 0.6, xlab = expression(varphi ~ "(predictor angle)"),
     ylab = expression(theta ~ "(response angle)"),
     xlim = c(-pi, pi), ylim = c(-pi, pi), axes = FALSE)
axis(1, at = c(-pi, -pi/2, 0, pi/2, pi),
     labels = expression(-pi, -pi/2, 0, pi/2, pi))
axis(2, at = c(-pi, -pi/2, 0, pi/2, pi),
     labels = expression(-pi, -pi/2, 0, pi/2, pi))
box()
lines(phig, 2 * atan(1.6 * sin(phig)), col = "gray60", lwd = 1.6, lty = 2)
lines(phig, mu_hat, col = "#c0392b", lwd = 2.6)
legend("topleft", bty = "n", lwd = c(2.6, 1.6), lty = c(1, 2),
       col = c("#c0392b", "gray60"),
       legend = c(expression(hat(mu)(varphi)), "truth"))
par(op)
```

Remember both axes are circles: the top and bottom edges are the same
line, as are the left and right — which is exactly what the torus shows
without the cut.

## The torus

Predictor angle $\varphi$ around the ring, response angle $\theta$ around
the tube; $(\varphi, \theta)$ maps to
$\big((R + r\cos\theta)\cos\varphi,\ (R + r\cos\theta)\sin\varphi,\
r\sin\theta\big)$. Base graphics is enough: `persp()` supplies the
projection matrix and everything is drawn through `trans3d()`, with
depth-based fading for the 3-D cue (the dashed circle is the outer
equator $\theta = 0$, the zero-response reference).

```{r torus, fig.height = 5.8}
torus_xyz <- function(phi, theta, R = 1.9, r = 0.95) {
  list(x = (R + r * cos(theta)) * cos(phi),
       y = (R + r * cos(theta)) * sin(phi),
       z = r * sin(theta))
}

# view-space depth from the persp transformation matrix (trans3d gives
# only the projected x/y; the z/w component orders front vs back)
depth3d <- function(x, y, z, pm) {
  p <- cbind(x, y, z, 1) %*% pm
  p[, 3] / p[, 4]
}

draw_torus <- function(dat, mu_hat, phig, R = 1.9, r = 0.95,
                       theta_view = 45, phi_view = 38) {
  op <- par(mar = c(0.2, 0.2, 0.2, 0.2))
  on.exit(par(op))
  lim <- R + r + 0.1
  pm <- persp(x = c(-lim, lim), y = c(-lim, lim),
              z = matrix(c(-r, -r, r, r), 2, 2),
              zlim = c(-r - 0.5, r + 0.5),
              theta = theta_view, phi = phi_view, d = 4,
              scale = FALSE, expand = 1,
              col = NA, border = NA, box = FALSE, axes = FALSE)

  ## wireframe: tube circles at fixed phi, ring circles at fixed theta
  thd <- seq(-pi, pi, length.out = 80)
  for (p in seq(-pi, pi, length.out = 49)[-1]) {
    w <- torus_xyz(rep(p, length(thd)), thd, R, r)
    lines(trans3d(w$x, w$y, w$z, pm), col = "gray88", lwd = 0.6)
  }
  phd <- seq(-pi, pi, length.out = 160)
  for (t in seq(-pi, pi, length.out = 25)[-1]) {
    w <- torus_xyz(phd, rep(t, length(phd)), R, r)
    lines(trans3d(w$x, w$y, w$z, pm), col = "gray88", lwd = 0.6)
  }

  ## outer equator (theta = 0): the zero-response reference line
  w <- torus_xyz(phd, rep(0, length(phd)), R, r)
  lines(trans3d(w$x, w$y, w$z, pm), col = "gray55", lty = 3, lwd = 0.9)

  ## data on the torus surface, depth-faded (near = opaque, far = faint)
  w <- torus_xyz(dat$phi, dat$theta, R, r)
  dp <- depth3d(w$x, w$y, w$z, pm)
  a <- 0.15 + 0.7 * (dp - min(dp)) / diff(range(dp))
  pt <- trans3d(w$x, w$y, w$z, pm)
  ord <- order(dp)  # paint far points first
  points(pt$x[ord], pt$y[ord], pch = 19,
         cex = 0.4 + 0.25 * a[ord],
         col = sapply(a[ord], function(ai) adjustcolor("#2c5aa0", ai)))

  ## fitted mean-direction curve mu-hat(phi), winding over the tube --
  ## painted in depth order, far half faint
  w <- torus_xyz(phig, mu_hat, R, r)
  dp <- depth3d(w$x, w$y, w$z, pm)
  cv <- trans3d(w$x, w$y, w$z, pm)
  a <- 0.25 + 0.75 * (dp - min(dp)) / diff(range(dp))
  seg <- data.frame(x0 = head(cv$x, -1), y0 = head(cv$y, -1),
                    x1 = tail(cv$x, -1), y1 = tail(cv$y, -1),
                    d = head(dp, -1), a = head(a, -1))
  seg <- seg[order(seg$d), ]
  segments(seg$x0, seg$y0, seg$x1, seg$y1,
           col = sapply(seg$a, function(ai) adjustcolor("#c0392b", ai)),
           lwd = 2.2 + 1.6 * seg$a, lend = 1)
  invisible(pm)
}

draw_torus(dat, mu_hat, phig)
```

The red curve is the fitted conditional mean direction
$\hat\mu(\varphi)$: a closed loop on the torus, rising over the tube
where the response leads the predictor and dipping below where it lags,
crossing the dashed equator at the two angles where $\mu = 0$.

## Same model, same numbers, in Python

This model class is the mgcv twin of
[pycircstat2](https://github.com/circstat/pycircstat2)'s
`CLRegression(..., family=vmlss)`; the two implementations are
differentially tested against each other on every release — identical
data, Newton-REML both sides — and agree to machine precision for
cyclic-spline models like this one (see `tests/testthat/test-vmlss-parity.R`
and `dev/parity/` in the repository).
