---
title: "Linear–circular regression on the cylinder"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5.4, dpi = 130
)
```

*One of three response–predictor geometries:
[C–L](circular-linear-regression.html) · **L–C** ·
[C–C](circular-circular-regression.html).*

Linear–circular (L–C) regression has an ordinary real-valued response
$y$ and a *circular* covariate $\varphi$ — energy demand vs time of day,
growth vs season, any "how does the level change around the cycle"
question.

An honest note first: **this case needs no `circlss` family at all.**
A circular response family is what `circlss` adds to mgcv; a circular
*covariate* is something mgcv already handles natively, with its cyclic
spline bases (`bs = "cc"` / `bs = "cp"`). The L–C model is a plain GAM —
this article completes the C–L / L–C / C–C trio and shows the pattern
(and the cylinder picture) precisely so you know where `circlss` is and
is not needed.

$$y_i \sim N\big(m(\varphi_i),\ \sigma^2(\varphi_i)\big),
\qquad m, \log\sigma \ \text{periodic in}\ \varphi,$$

fit distributionally with mgcv's own `gaulss()` — structurally the
exact sibling of a `vmlss`/`pnlss` call, two formulas and cyclic
smooths, just with a Gaussian response.

## Simulate and fit

```{r}
library(mgcv)

set.seed(20260612)
n <- 300
phi <- runif(n, -pi, pi)
m_true <- 2 + 1.5 * sin(phi) + 0.8 * cos(2 * phi)
sd_true <- exp(-0.5 + 0.6 * cos(phi))
dat <- data.frame(y = rnorm(n, m_true, sd_true), phi = phi)

b <- gam(list(y ~ s(phi, bs = "cc", k = 10),
                ~ s(phi, bs = "cc", k = 10)),
         family = gaulss(), data = dat, method = "REML",
         knots = list(phi = c(-pi, pi)))
summary(b)
```

(`gaulss` models the mean and the *precision* $\tau \approx 1/\sigma$;
its second fitted column is $\tau$, so $\hat\sigma = 1/\hat\tau$.)

## Flat view

Mean curve with its 95% band, plus the heteroscedasticity readout
$\hat\sigma(\varphi)$ — both periodic, closing seamlessly at $\pm\pi$:

```{r flat-lc, fig.height = 4.2}
phig <- seq(-pi, pi, length.out = 400)
pr <- predict(b, newdata = data.frame(phi = phig), se.fit = TRUE)
m_hat <- pr$fit[, 1]
m_lo <- m_hat - 1.96 * pr$se.fit[, 1]
m_hi <- m_hat + 1.96 * pr$se.fit[, 1]
sd_hat <- 1 / predict(b, newdata = data.frame(phi = phig),
                      type = "response")[, 2]

op <- par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 1))
plot(dat$phi, dat$y, pch = 19, col = adjustcolor("#2c5aa0", 0.45),
     cex = 0.6, xlab = expression(varphi), ylab = "y",
     main = "mean", axes = FALSE)
axis(1, at = c(-pi, 0, pi), labels = expression(-pi, 0, pi))
axis(2)
box()
polygon(c(phig, rev(phig)), c(m_lo, rev(m_hi)), border = NA,
        col = adjustcolor("#c0392b", 0.15))
lines(phig, 2 + 1.5 * sin(phig) + 0.8 * cos(2 * phig),
      col = "gray60", lwd = 1.6, lty = 2)
lines(phig, m_hat, col = "#c0392b", lwd = 2.4)

plot(phig, sd_hat, type = "l", col = "#c0392b", lwd = 2.4,
     xlab = expression(varphi), ylab = expression(hat(sigma)(varphi)),
     main = "scale", axes = FALSE, ylim = range(0, sd_hat, exp(0.1)))
axis(1, at = c(-pi, 0, pi), labels = expression(-pi, 0, pi))
axis(2)
box()
lines(phig, exp(-0.5 + 0.6 * cos(phig)), col = "gray60",
      lwd = 1.6, lty = 2)
par(op)
```

## The cylinder

For L–C the cylinder stands upright: the circular covariate runs around
the can, the response is the height. $(\varphi, y)$ maps to
$\big(r\cos\varphi,\ r\sin\varphi,\ h(y)\big)$, and the fitted mean is a
closed loop around the can at varying height, with its ribbon. Same
toolkit as the [torus](circular-circular-regression.html): `persp()`
for the projection, `trans3d()` for drawing, depth-fading for the 3-D
cue (the dashed ring sits at the overall mean height):

```{r can-helpers}
# view-space depth from the persp transformation matrix
depth3d <- function(x, y, z, pm) {
  p <- cbind(x, y, z, 1) %*% pm
  p[, 3] / p[, 4]
}

draw_can <- function(dat, m_hat, phig, lo = NULL, hi = NULL,
                     r = 1, H = 1.5, theta_view = 30, phi_view = 16) {
  yl <- range(dat$y, lo, hi)
  can_xyz <- function(phi, y) {
    h <- (y - yl[1]) / diff(yl) * 2 * H - H
    list(x = r * cos(phi), y = r * sin(phi), z = h)
  }
  op <- par(mar = c(0.2, 0.2, 0.2, 0.2))
  on.exit(par(op))
  pm <- persp(x = c(-r - 0.3, r + 0.3), y = c(-r - 0.3, r + 0.3),
              z = matrix(c(-H, -H, H, H), 2, 2),
              zlim = c(-H - 0.3, H + 0.3),
              theta = theta_view, phi = phi_view, d = 4,
              scale = FALSE, expand = 1,
              col = NA, border = NA, box = FALSE, axes = FALSE)

  ## wireframe: vertical generators + horizontal rings
  hd <- seq(yl[1], yl[2], length.out = 40)
  for (p in seq(-pi, pi, length.out = 49)[-1]) {
    w <- can_xyz(rep(p, length(hd)), hd)
    lines(trans3d(w$x, w$y, w$z, pm), col = "gray88", lwd = 0.6)
  }
  phd <- seq(-pi, pi, length.out = 160)
  for (t in seq(yl[1], yl[2], length.out = 13)) {
    w <- can_xyz(phd, rep(t, length(phd)))
    lines(trans3d(w$x, w$y, w$z, pm), col = "gray88", lwd = 0.6)
  }

  ## reference ring at the overall mean height
  w <- can_xyz(phd, rep(mean(dat$y), length(phd)))
  lines(trans3d(w$x, w$y, w$z, pm), col = "gray55", lty = 3, lwd = 0.9)

  ## data on the surface, depth-faded
  w <- can_xyz(dat$phi, dat$y)
  dp <- depth3d(w$x, w$y, w$z, pm)
  a <- 0.15 + 0.7 * (dp - min(dp)) / diff(range(dp))
  pt <- trans3d(w$x, w$y, w$z, pm)
  ord <- order(dp)
  points(pt$x[ord], pt$y[ord], pch = 19,
         cex = 0.4 + 0.25 * a[ord],
         col = sapply(a[ord], function(ai) adjustcolor("#2c5aa0", ai)))

  ## 95% ribbon, depth-ordered translucent quads
  if (!is.null(lo) && !is.null(hi)) {
    K <- 4
    i0 <- seq_len(length(phig) - 1)
    quads <- list()
    for (k in seq_len(K)) {
      t0 <- lo + (hi - lo) * (k - 1) / K
      t1 <- lo + (hi - lo) * k / K
      for (i in i0) {
        hh <- c(t0[i], t0[i + 1], t1[i + 1], t1[i])
        pp <- c(phig[i], phig[i + 1], phig[i + 1], phig[i])
        w <- can_xyz(pp, hh)
        quads[[length(quads) + 1]] <-
          list(p = trans3d(w$x, w$y, w$z, pm),
               d = mean(depth3d(w$x, w$y, w$z, pm)))
      }
    }
    dq <- vapply(quads, `[[`, numeric(1), "d")
    aq <- 0.05 + 0.13 * (dq - min(dq)) / diff(range(dq))
    for (j in order(dq)) {
      polygon(quads[[j]]$p$x, quads[[j]]$p$y, border = NA,
              col = adjustcolor("#c0392b", aq[j]))
    }
  }

  ## fitted mean curve: a closed loop around the can, depth-painted
  w <- can_xyz(phig, m_hat)
  dp <- depth3d(w$x, w$y, w$z, pm)
  cv <- trans3d(w$x, w$y, w$z, pm)
  a <- 0.25 + 0.75 * (dp - min(dp)) / diff(range(dp))
  seg <- data.frame(x0 = head(cv$x, -1), y0 = head(cv$y, -1),
                    x1 = tail(cv$x, -1), y1 = tail(cv$y, -1),
                    d = head(dp, -1), a = head(a, -1))
  seg <- seg[order(seg$d), ]
  segments(seg$x0, seg$y0, seg$x1, seg$y1,
           col = sapply(seg$a, function(ai) adjustcolor("#c0392b", ai)),
           lwd = 2.2 + 1.6 * seg$a, lend = 1)
  invisible(pm)
}
```

```{r can-lc, fig.height = 5.6}
draw_can(dat, m_hat, phig, lo = m_lo, hi = m_hi)
```

The loop closes on itself — that is the cyclic basis doing its job; an
unconstrained `s(phi)` would let the two ends of the cycle disagree.

## Where circlss picks up

Flip the roles — circular *response* — and you leave what plain mgcv
ships: that is exactly the gap `circlss` fills, with `vmlss()` and
`pnlss()` general families whose every parameter gets its own smooth.
See the [C–L](circular-linear-regression.html) article for the linear
covariate case (and the helix that needs `pnlss`), and the
[C–C](circular-circular-regression.html) article for the doubly
circular case on the torus.
